Late Stage Review

### 2020-01-23 papers/lsr/lsr.tex
LSR

\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWab}{\mathbb{W}^{(a,b)}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}
\newcommand{\bigQ}{\mathbb{Q}}
\newcommand{\bigQab}{\bigQ^{(a,b)}}

@@

@@

\section{Orthogonal polynomials on the circle arc}

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the Lanczos algorithm as implemented in the ApproxFun package \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to \cite{huybrechs2010fourier}:
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In the work in this section we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations. For now, we have focused on constant Dirichlet boundary conditions, as well as zero Dirichlet and Neumann boundary conditions, with other types of boundary conditions requiring future work.

\subsection{Explicit construction}

Here, we have developed sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums. More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\} \label{eqn:domaindef}
\end{align}
where either of the following conditions hold:

@@

For simplicity of presentation we focus on the disk-slice, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$, and discuss an extension to other the end-disk-slice in the appendix (note we omit detailed discussion  on the extension to trapezium in this document).

@@

$\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
assuming that either \condref{trap} or \condref{disk} with  \(\jacw^{(a,b)}\) being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$) hold.
That is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}

@@

%Again, for clarity we will use the notation that for $p : \Omega \to \R$, $\norm{p}_{\Wab}^2 := \ip<p, p>_{\Wab}$.

@@

%\begin{proof}
%First, note that:
%	\genjacw^{(a,b,c) \: \prime}(x) &= - a \: \genjacw^{(a-1,b,c)}(x) + b \: \genjacw^{(a,b-1,c)}(x) + c \: \rho(x) \: \rho'(x) \:\genjacw^{(a,b,c-2)}(x), \label{eqn:derivativeofweightgenjac} \\
%	\jacw^{(c) \: \prime}(y) &= - 2c \: y \: \jacw^{(c-1)}(y), \label{eqn:derivativeofweightjac} \\
%	\rho(x) \: \rho'(x) &= -x. \label{eqn:rhoderivative}
%We proceed with the case for the operator $D_y^{(a,b,c)}$ for partial differentiation by $y$. Since $\{\hdop^{(a,b,c+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkabc = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^y \: \hdop^{(a, b, c+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
%	c_{m,j}^y = {\ip<\pddy \hdopnkabc, \hdop^{(a, b, c+1)}_{m,j} >_{W^{(a, b, c+1)}}}{\norm{\hdop^{(a, b, c+1)}_{m,j}}^{-2}_{W^{(a, b+1)}}}.
%Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
%	&\ip<\pddy \hdopnkabc, \hdop^{(a, b, c+1)}_{m,j} >_{W^{(a, b, c+1)}} \\
%	&= \iint_\Omega \: \Big[ \genjacnmk^{(a,b,2c+2k+1)}(x) \: \rho(x)^{k-1} \: \jac_k^{(c,c) \: \prime}\fpr(\frac{y}{\rho(x)}) \\
%	& \quad \quad \quad \quad \quad \quad \cdot \: \genjacmmj^{(a,b,2c+2j+3)}(x) \: \rho(x)^{j} \: \jac_j^{(c+1,c+1)}\fpr(\frac{y}{\rho(x)}) \Big] \: \D y \: \D x \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \rho(x)^{j-k+1} \: \genjacmmj^{(a,b,2c+2j+3)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(c+1)} \: \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c+1)}}
%Now, using (\ref{eqn:derivativeofweightjac}), integration-by-parts, and noting that the weight $\jacw^{(c)}$ is a polynomial of degree $2c$ and vanishes at the limits of the integral for positive parameter $c$, we have that
%	\normjac^{(c+1)} \: \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c+1)}} &= \int_{\gamma}^\delta \: \jac_k^{(c,c) \: \prime}(y) \: \jac_{j}^{(c+1,c+1)}(y) \: \jacw^{(c+1)}(y) \: \D y \\
%	&= - \int_{-1}^1 \: \jac_k^{(c,c)}(y) \: \ddy [\jacw^{(c+1)}(y) \: \jac_{j}^{(c+1,c+1)}(y)] \: \D y \\
%	&= - \int_{-1}^1 \: \jac_k^{(c,c)} \: [ \jac_{j}^{(c+1,c+1) \: \prime} \: \jacw^{(c+1)} - 2 c \: y \: \jac_{j}^{(c+1,c+1)} \:\jacw^{(c)} ] \: \D y \\
%	&= - \: \normjac^{(c)} \: \ip< \jac_k^{(c,c)}, \: \jacw^{(1)} \: \jac_{j}^{(c+1,c+1) \: \prime} -2c \:y \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c)}}
%which is zero for $j < k-1$ by orthogonality. Further, when $j = k-1$, we have that
%	&\normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \rho(x)^{j-k+1} \: \genjacmmj^{(a,b,2c+2j+3)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \genjacmmj^{(a,b,2c+2k+1)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \delta_{n,m+1},
%showing that the only possible non-zero coefficient is when $m=n-1, j=k-1$. Finally,
%	c_{n-1,k-1}^y &= \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{k-1}^{(c+1,c+1)} >_{\jacw^{(c+1)}}.
%We next consider the case for the operator $D_x^{(a,b,c)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1, c+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkabc = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^x \: \hdop^{(a+1, b+1, c+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. As before, we can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2, n-3$, $j = k, k-1, k-2$ and $0 \le j \le m$. First, note that
%	c_{m,j}^x = {\ip<\pddx \hdopnkabc, \hdop^{(a+1, b+1, c+1)}_{m,j} >_{W^{(a+1, b+1, c+1)}}}{\norm{\hdop^{(a+1, b+1, c+1)}_{m,j}}^{-2}_{W^{(a+1, b+1, c+1)}}}.
%Now, again using the change of variable $t= \frac{y}{\rho(x)}$, we have that
%	&\ip<\pddx \hdopnkabc, \hdop^{(a+1, b+1, c+1)}_{m,j} >_{W^{(a+1, b+1, c+1)}} \nonumber \\ 
%	&= \Big( \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1) \: \prime} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j+1} \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \Big) \nonumber \\
%	&\quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \jac_k^{(c,c)} \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \: \D t \Big) \nonumber \\ 
%	& \quad + k \: \Big( \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j} \: \rho' \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \Big) \nonumber \\
%	& \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta \jac_k^{(c,c)} \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \: \D t \Big) \nonumber \\
%	& \quad - \Big( \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j} \: \rho' \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \Big) \nonumber \\
%	& \quad \quad \quad \quad \quad \cdot \: \Big( \int_\gamma^\delta t \: \jac_k^{(c,c) \: \prime} \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \: \D t \Big). \label{eqn:dxsparsity}
%We will first show that the second factor of each term in (\ref{eqn:dxsparsity}) are zero for $j < k - 2$ and also for $j = k - 1$. To this end, observe that, for any integer $c$, $\jac^{(c,c)}(-t) = (-1)^k \: \jac^{(c,c)}(t)$ and so $\jac_k^{(c,c)}$ is an even polynomial for even $k$, and an odd polynomial for odd $k$. Thus, $\jac_k^{(c,c)} \: \jac_{k-1}^{(c+1,c+1)}$ is an odd polynomial for any $k$. Hence
%	\int_\gamma^\delta \jac_k^{(c,c)} \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \: \D t = \int_{-\delta}^\delta \jac_k^{(c,c)} \: \jac_j^{(c+1,c+1)} \: \jacw^{(1)} \: \jacw^{(c)} \: \D t 
%is zero for $j < k - 2$ by orthogonality, and is zero for $j = k-1$ due to symmetry over the domain. Moreover, $t \: \jac_k^{(c,c) \: \prime}(t) \: \jac_j^{(c+1,c+1)}(t)$ is also an odd polynomial for any $k$ and so
%	\int_\gamma^\delta t \: \jac_k^{(c,c) \: \prime}(t) \: \jac_j^{(c+1,c+1)}(t) \: \jacw^{(c+1)}(t) \: \D t
%is zero for $j = k-1$ due to symmetry over the domain, and
%	&\int_\gamma^\delta t \: \jac_k^{(c,c) \: \prime} \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \: \D t \\
%	&= - \int_{-\delta}^\delta \jac_k^{(c,c)} \: \frac{\D}{\D t} \big[ t \: \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} \big] \: \D t \\
%	&= - \int_{-\delta}^\delta \jac_k^{(c,c)} \: \big[ \jac_j^{(c+1,c+1)} \: \jacw^{(c+1)} + t \: \jac_j^{(c+1,c+1) \: \prime} \: \jacw^{(c+1)} - 2 c \: t^2 \: \jac_j^{(c+1,c+1)} \: \jacw^{(c)} \big] \: \D t \\
%	&= - \: \normjac^{(c)} \: \ip< \jac_k^{(c,c)}, \: \jac_j^{(c+1,c+1)} \: \jacw^{(1)} + t \: \jac_{j}^{(c+1,c+1) \: \prime} \: \jacw^{(1)} - 2c \: t^2 \: \jac_{j}^{(c+1,c+1)}  >_{\jacw^{(c)}}
%which is zero for $j < k - 2$ by orthogonality. Thus, (\ref{eqn:dxsparsity}) is zero for $j \notin \{k-2, k\}$.
%Now, using (\ref{eqn:derivativeofweightgenjac}), integration-by-parts, and noting that the weight $\genjacw^{(a,b,2c)}$ is a polynomial degree $a+b+2c$ and vanishes at the limits of the integral for positive parameters $a,b,c$, we have that
%	&\int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1) \: \prime} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j+1} \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \nonumber  \\
%	&= \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1) \: \prime} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \genjacw^{(a+1, b+1, 2c+k+j+3)} \: \D x \nonumber \\
%	&= - \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \ddx \Big[ \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \genjacw^{(a+1, b+1, 2c+k+j+3)} \Big] \: \D x \nonumber \\
%	&= - \int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \Big\{ \genjacmmj^{(a+1, b+1, 2c+2j+3) \: \prime} \: \genjacw^{(a+1, b+1, 2c+k+j+3)} \nonumber \\
%	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (2c+k+j+3) \: \rho \: \rho' \:  \genjacw^{(a+1, b+1, 2c+k+j+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \nonumber \\
%	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (b+1) \: \genjacw^{(a+1, b, 2c+k+j+3)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \nonumber \\
%	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad - (a+1) \: \genjacw^{(a, b+1, 2c+k+j+3)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \Big\} \: \D x \nonumber \\
%	&= - \: \normgenjac^{(a, b, 2c+2k+1)} \:  \Big\{ \ip<\genjacnmk^{(a, b, 2c+2k+1)}, \: \genjacw^{(1, 1, j-k+2)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3) \: \prime}>_{\genjacw^{(a, b, 2c+2k+1)}} \nonumber \\
%		&\quad \quad \quad + (2c+k+j+3) \: \ip<\genjacnmk^{(a, b, 2c+2k+1)}, \: \rho \: \rho' \: \genjacw^{(1,1, j-k)} \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}} \nonumber \\
%		&\quad \quad \quad + (b+1) \: \ip<\genjacnmk^{(a, b, 2c+2k+1)}, \: \genjacw^{(1,0, j-k+2)} \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}} \nonumber \\
%		&\quad \quad \quad - (a+1) \: \ip<\genjacnmk^{(a, b, 2c+2k+1)}, \: \genjacw^{(0,1, j-k+2)} \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}} \Big\}. \label{eqn:dxsparsityR}
%By recalling (\ref{eqn:rhoderivative}) and noting that $j-k$ is even by the earlier argument, we can see $\rho \: \rho' \: \genjacw^{(1,1, j-k)}$, $\genjacw^{(1,0, j-k+2)}$ and $\genjacw^{(1,0, j-k+2)}$ are all polynomials, and further that 
%$$\text{deg}(\rho \: \rho' \: \genjacw^{(1,1, j-k)}) = \text{deg} (\genjacw^{(1,0, j-k+2)}) = \text{deg} (\genjacw^{(0,1, j-k+2)}) = 3 + j-k.$$ 
%Hence, by orthogonality, each term in (\ref{eqn:dxsparsityR}) is is zero for $m-j+3+j-k < n-k \iff m < n - 3$.
%Finally,
%	&\int_\alpha^\beta \: \genjacnmk^{(a, b, 2c+2k+1)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} \: \rho^{k+j} \: \rho' \: \genjacw^{(a+1, b+1, 2c+2)} \: \D x \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a, b, 2c+2k+1)}, \: \rho \: \rho' \: \genjacw^{(1, 1, j-k)} \: \genjacmmj^{(a+1, b+1, 2c+2j+3)} >_{\genjacw^{(a, b, 2c+2k+1)}}
%which is also zero for $m < n - 3$. Thus
%	&\ip<\pddx \hdopnkabc, \hdop^{(a+1, b+1, c+1)}_{m,j} >_{W^{(a+1, b+1, c+1)}} = 0 \qfor m < n - 3, \: j \notin \{k-2, k\},
%showing that the only possible non-zero coefficients $c_{m,j}^x$ are when $m = n-3,\dots,n$ and $j = k-2,k$ $(j \le m)$.
%We can gain the non-zero entries of the weighted differential operators similarly, by noting that for the disk-slice
%	\pddx \Wabc(x,y) &= -a W^{(a-1, b, c)}(x,y) + bW^{(a, b, c)}(x,y) + 2c \rho(x) \: \rho'(x) \: W^{(a, b, c-1)} \label{eqn:weightderivativex} \\
%	\pddy \Wabc(x,y) &= -2c \: y \: W^{(a, b, c-1)}(x,y) \label{eqn:weightderivativey}
%and also that
%	\ip< \Wabc \hdopnkabc, W^{(\tilde{a},\tilde{b},\tilde{c})} \hdopmj^{(\tilde{a},\tilde{b},\tilde{c})} >_{W^{(-\tilde{a},-\tilde{b},-\tilde{c})}} = \ip< \hdopnkab, \hdopmj^{(\tilde{a},\tilde{b})} >_\Wabc.
%\end{proof}

@@

%\begin{proof}
%We proceed with the case for the non-weighted operators $T^{(a,b)\to(a+\tilde{a},b+\tilde{b},c+\tilde{c})}$, where $\tilde{a}, \tilde{b}, \tilde{c} \in \{0,1\}$. Since $\{\hdop^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}_{m,j}\}$ for $m = 0,\dots,n$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n$ polynomial, we can expand $\hdopnkabc = \sum_{m=0}^{n} \sum_{j=0}^m c_{m,j} \: \hdop^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We will show that the only non-zero coefficients are for $m \ge n - \tilde{a} - \tilde{b} - 2\tilde{c} $, $j \ge k-2\tilde{c}$ and $0 \le j \le m$. First, note that
%\begin{align*}
%	c_{m,j} = {\ip< \hdopnkabc, \hdop^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}_{m,j} >_{W^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}}}{\norm{\hdop^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}_{m,j}}^{-2}_{W^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}}}.
%\end{align*}
%Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
%\begin{align*}
%	&\ip< \hdopnkabc, \hdop^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}_{m,j} >_{W^{(a+\tilde{a},b+\tilde{b},c+\tilde{c})}} \\
%	&\quad = \normgenjac^{(a+\tilde{a},b+\tilde{b},2c+2\tilde{c})} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \rho(x)^{k+j+1} \: \genjacmmj^{(a+\tilde{a},b+\tilde{b},2c+2\tilde{c}+2j+1)} >_{\genjacw^{(a+\tilde{a},b+\tilde{b},2c+2\tilde{c})}} \\ 
%	&\quad \quad \quad \cdot \: \normjac^{(c+\tilde{c})} \: \ip< \jac_k^{(c,c)}, \: \jac_{j}^{(c+\tilde{c},c+\tilde{c})} >_{\jacw^{(c+\tilde{c})}} \\
%	&\quad= \normgenjac^{(a,b,2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \genjacw^{(\tilde{a}, \tilde{b}, \: 2\tilde{c}+j-k)} \: \genjacmmj^{(a+\tilde{a},b+\tilde{b},2c+2\tilde{c}+2j+1)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&\quad \quad \quad \cdot \: \normjac^{(c)} \: \ip< \jac_k^{(c,c)}, \: \jacw^{(\tilde{c})} \: \jac_{j}^{(c+\tilde{c},c+\tilde{c})} >_{\jacw^{(c)}}.
%\end{align*}
%Since $\jacw^{(\tilde{c})}$ is a polynomial degree $2\tilde{c}$, we have that the above is then zero for $j < k-2\tilde{c}$. Further, since $\genjacw^{(\tilde{a}, \tilde{b}, \: 2\tilde{c}+j-k)}$ is a polynomial of degree $\tilde{a}+\tilde{b}+2\tilde{c}+j-k$, we have that the above is zero for $m-j+\tilde{a}+\tilde{b}+2\tilde{c}+j-k < n-k \iff m < n - \tilde{a}-\tilde{b}-2\tilde{c}$.
%
%The sparsity argument for the weighted parameter transformation operators follows similarly.
%\end{proof}

@@

%To obtain the recurrence coefficients for the $\{\genjac_n^{(a,b,c)}\}$ OPs in (\ref{eqn:Hrecurrence}), we use a variant of the Stieltjes procedure \cite{gautschi1982generating} where the polynomials are expressed as Chebyshev polynomial expansions and the inner products are calculated via Clenshaw--Curtis quadrature. This has the benefit that it is easier to incorporate high-precision arithmetic, which we use to overcome ill-conditioning present when $c$ is large, as required for large $n$ in \eqref{eq:diskpolys}. The ApproxFun.jl \cite{ApproxFun} package gives a convenient way to manipulate Chebyshev and Jacobi expansions, which we use to calculate the inner products and norms in this algorithm, utilising the {\tt BigFloat} type to handle high-precision calculations.

>>>>

\newcommand{\element}{\tau}
\newcommand{\refelement}{\hat{\tau}}
\newcommand{\FEset}{\mathcal{T}}
\newcommand{\bigW}{\mathbb{W}}
\newcommand{\bigWab}{\mathbb{W}^{(a,b)}}
\newcommand{\bigWii}{{\mathbb{W}^{(1,1)}}}
\newcommand{\bigQ}{\mathbb{Q}}
\newcommand{\bigQab}{\bigQ^{(a,b)}}
\newcommand{\Qab}{Q^{(a,b)}}

@@

\newcommand{\xvec}{\mathbf{x}}

@@

@@

Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations. In this paper we focus on constant Dirichlet boundary conditions, as well as zero Dirichlet and Neumann boundary conditions, with other types of boundary conditions requiring future work.

\subsection{Explicit construction}
Here, we have developed sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums. More precisely, we consider the solution of partial differential equations on the domain
\begin{align}\label{eqn:domaindef}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
where either of the following conditions hold:

@@

For simplicity of presentation we focus on the disk-slice, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$, and discuss an extension to the end-disk-slice in the appendix (note we omit detailed discussion on the extension to the trapezium in this document).

@@

The OPs $\{\hdopnk^{(a,b,c,d)}\}$ are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: \jacw^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
assuming that either \condref{trap} or \condref{disk} with  \(\jacw^{(a,b)}\) being an even function (i.e. $a = b$, and we can hence denote the weight as $\jacw^{(a)}(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$) hold.
That is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}

@@

@@

@@

@@

It is possible to obtain the recurrence coefficients for the $\{\genjac_n^{(a,b,c)}\}$ OPs in (\ref{eqn:Hrecurrence}), by careful application of the Christoffel--Darboux formula \cite[18.2.12]{DLMF}. We explain the process here for the disk-slice case, however we note that a similar but simpler argument holds for the trapezium case. We thus first need to define a new set of `interim' 1D OPs.
\begin{definition}\label{def:InterimOPconstruction}
Let $\genjactw^{(a,b,c,d)}(x) := (\beta - x)^a \: (x - \alpha)^{b} \: (1-x)^{c} \: (1+x)^{d} $ be a weight function on the interval $(\alpha, \beta)$, and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjactw^{(a,b,c,d)}} &:= \frac{1}{\normgenjact^{(a,b,c,d)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjactw^{(a,b,c,d)}(x) \: \D x \label{eqn:ipgenjact}
\end{align}
where
\begin{align}
	\normgenjact^{(a,b,c,d)} := \int_\alpha^\beta \: \genjactw^{(a,b,c,d)}(x) \: \D x
\end{align}
Denote the four-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjact_n^{(a,b,c,d)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjact}).
\end{definition}
Note that the OPs $\{\genjac_n^{(a,b,2c)}\}$ are then equivalent to the OPs $\{\genjact_n^{(a,b,c,c)}\}$. Let the recurrence coefficients for the OPs $\{\genjact_n^{(a,b,c,d)}\}$ be given by:
\begin{align}
	x \: \genjact_n^{(a,b,c,d)}(x) = \tilde{\beta}_n^{(a,b,c,d)} \: \genjact_{n+1}^{(a,b,c,d)}(x) + \tilde{\alpha}_n^{(a,b,c,d)} \:\genjact_n^{(a,b,c,d)}(x) + \tilde{\beta}_{n-1}^{(a,b,c,d)} \: \genjact_{n-1}^{(a,b,c,d)}(x)
\end{align}
\begin{proposition}
There exist constants $\mathcal{C}_n^{(a,b,c,d)}$, $\mathcal{D}_n^{(a,b,c,d)}$ such that
\begin{align}
	\genjact_n^{(a,b,c+1,d)}(x) &= \mathcal{C}_n^{(a,b,c,d)} \: \sum_{k=0}^n \: \genjact_{k}^{(a,b,c,d)}(1) \: \genjact_{k}^{(a,b,c,d)}(x) \label{eqn:genjacexpansioni} \\
	\genjact_n^{(a,b,c,d+1)}(x) &= \mathcal{D}_n^{(a,b,c,d)} \: \sum_{k=0}^n \: \genjact_{k}^{(a,b,c,d)}(-1) \: \genjact_{k}^{(a,b,c,d)}(x) \label{eqn:genjacexpansionii}
\end{align}
\end{proposition}
